---
title: "Final_code"
author: "Sonika Shrestha"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Data Exploration

```{r}

#install.packages("ROCR")
#install.packages("rocc")
#install.packages("ROCit")
#install.packages("pROC")
#install.packages("caTools")
#install.packages("precrec")
#install.packages("AUC")
#install.packages("verification")

library(ggplot2)
library(corrplot)
library(VIM)
library(imputeTS)

# Logistic Regression Performance
library(ROCR)
library(rocc)
library(ROCit)
library(pROC)
library(caTools)
library(precrec)
library(AUC)
library(verification)
library(caret)

```

### Load the dataset in R Studio. Examine the first few rows of data using R. Explain your findings. Did you notice anything abnormal or interesting?

```{r}
data <-read.csv("patients.csv")

```

###Provide summary statistics. Calculate the mean, median, standard deviation, and quartiles for each independent variable. Explain your results.

```{r}
# Calculate summary statistics
summary_stats <- function(x) {
  mean_val <- mean(x)
  median_val <- median(x)
  sd_val <- sd(x)
  quartiles <- quantile(x, probs = c(0.25, 0.5, 0.75))
  
  return(list(
    mean = mean_val,
    median = median_val,
    sd = sd_val,
    quartiles = quartiles
  ))
}

# Apply the summary_stats function to each variable
result <- lapply(data, summary_stats)

# Create a data frame with the results
result_df <- data.frame(
  Variable = names(result),
  Mean = sapply(result, function(x) x$mean),
  Median = sapply(result, function(x) x$median),
  SD = sapply(result, function(x) x$sd),
  Q1 = sapply(result, function(x) x$quartiles[1]),
  Q2 = sapply(result, function(x) x$quartiles[2]),
  Q3 = sapply(result, function(x) x$quartiles[3])
)

# Load the required packages
library(knitr)

# Print the table
kable(result_df)
```

### Using the ggplot2 library, create any five visualizations. Explain your reasoning for selecting those visualizations. Explain the output of each visualization. What are the insights your visualizations reveal about the dataset? 


##Explain Findings:
### Find missing values for each independent variable and fill them with median values. The missing values for independent variables in the dataset are coded 0.
```{r}
# Given that the missing values for variables are coded 0

# Replace zeros with NA in the columns except the dependent column

columns_to_exclude <- c(9)

data[, -columns_to_exclude][data[, -columns_to_exclude] == 0] <- NA

summary(data)

independent_vars <- c("Pregnancies","Glucose" , "BloodPressure", "SkinThickness", "Insulin", "BMI" , "Pedigree", "Age")

med_data <- data
# Replacing NA values with median for each independent variable
for (var in independent_vars){
  # Calculate the median of the variable, ignoring NA values
  med_value<- median(med_data[, var], na.rm=TRUE)
  med_value
#Replace NA values with the median in the new dataset
  med_data[is.na(med_data[, var]), var] <- med_value
}
```

```{r}
summary(med_data)
```

### Find outliers for each independent variable using the IQR rule

```{r}
# Function to find outliers using the IQR rule
find_outliers <- function(x) {
  q <- quantile(x, probs = c(0.25, 0.75))
  iqr <- q[2] - q[1]
  lower_bound <- q[1] - 1.5 * iqr
  upper_bound <- q[2] + 1.5 * iqr
  outliers <- x[x < lower_bound | x > upper_bound]
  return(outliers)
}

# Apply the find_outliers function to each variable in your original dataset
outliers <- lapply(med_data, find_outliers)


# Create a data frame with the outliers
outliers_df <- data.frame(
  Variable = names(outliers),
  Outliers = sapply(outliers, function(x) paste(x, collapse = ", "))
)

# Load the required packages
library(knitr)

# Print the table
kable(outliers_df)
```

```{r}
#replace outliers

independent_vars <- c("Pregnancies","Glucose" , "BloodPressure", "SkinThickness", "Insulin", "BMI" , "Pedigree", "Age")


for (rows in independent_vars) {
  q1 <- quantile(med_data[[rows]], 0.25, na.rm = TRUE)
  q3 <- quantile(med_data[[rows]], 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  
  # Perform winsorization on the specific variable
  med_data[[rows]] <- ifelse(med_data[[rows]] < lower_bound, lower_bound,
                             ifelse(med_data[[rows]] > upper_bound, upper_bound, med_data[[rows]]))
}

summary(med_data)

```

```{r}

# Function to find outliers using the IQR rule
find_outliers <- function(x) {
  q <- quantile(x, probs = c(0.25, 0.75))
  iqr <- q[2] - q[1]
  lower_bound <- q[1] - 1.5 * iqr
  upper_bound <- q[2] + 1.5 * iqr
  outliers <- x[x < lower_bound | x > upper_bound]
  return(outliers)
}

# Apply the find_outliers function to each variable in your original dataset
outliers <- lapply(med_data, find_outliers)


# Create a data frame with the outliers
outliers_df <- data.frame(
  Variable = names(outliers),
  Outliers = sapply(outliers, function(x) paste(x, collapse = ", "))
)

# Load the required packages
library(knitr)

# Print the table
kable(outliers_df)

```

### After dealing with Outliers

```{r}

library(ggplot2)
library(gridExtra)

# Box plot for Glucose values when Diagnosis is 0

p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = Glucose, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "Diagnosis", y = "Glucose") +
  ggtitle("Glucose values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for Glucose values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = Glucose, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "Glucose", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: Glucose values vs Diagnosis")



# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2)  

```

```{r}


p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = BloodPressure, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "BloodPressure", y = "BloodPressure") +
  ggtitle("BloodPressure values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for BloodPressure values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = BloodPressure, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "BloodPressure", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: BloodPressure values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 


```



```{r}

p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = SkinThickness, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "SkinThickness", y = "SkinThickness") +
  ggtitle("SkinThickness values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for SkinThickness values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = SkinThickness, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "SkinThickness", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: SkinThickness values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 

```



```{r}


p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = Insulin, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "Insulin", y = "Insulin") +
  ggtitle("Insulin values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for Insulin values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = Insulin, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "Insulin", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: Insulin values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 

```

```{r}


p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = BMI, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "BMI", y = "BMI") +
  ggtitle("BMI values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for BMI values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = BMI, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "BMI", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: BMI values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 

```

```{r}


p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = Pedigree, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "Pedigree", y = "Pedigree") +
  ggtitle("Pedigree values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for Pedigree values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = Pedigree, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "Pedigree", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: Pedigree values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 


```

```{r}

p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = Age, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "Age", y = "Age") +
  ggtitle("Age values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for Age values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = Age, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "Age", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: Age values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 


```

```{r}


p1 <- ggplot(med_data, aes(x = factor(Diagnosis), y = Pregnancies, fill = factor(Diagnosis))) +
  geom_boxplot() +
  labs(x = "Pregnancies", y = "Pregnancies") +
  ggtitle("Pregnancies values vs Diagnosis") +theme(legend.position = "bottom", legend.box = "horizontal")+
  scale_fill_manual(values = c("0" = "blue", "1" = "red"))

# Density plot for Pregnancies values when Diagnosis is 0
p3 <- ggplot(med_data, aes(x = Pregnancies, fill = factor(Diagnosis))) +
  geom_density(alpha = 0.5) +
  labs(x = "Pregnancies", y = "Density") +theme(legend.position = "bottom", legend.box = "horizontal")+
  ggtitle("Density Plot: Pregnancies values vs Diagnosis")

# Arrange the box plots and density plots in two separate frames
grid.arrange(p1, p3, ncol = 2) 

```

```{r}
# Create a histogram
hist(med_data$Pregnancies, main = "Distribution of Pregnancies", xlab = "No. of pregnancies", ylab = "Frequency")

hist(med_data$Glucose, main = "Distribution of Glucose", xlab = "Glucose amount", ylab = "Frequency")

hist(med_data$BloodPressure, main = "Distribution of BP", xlab = "Blood Pressure", ylab = "Frequency")

hist(med_data$SkinThickness, main = "Distribution of Skin Thickness", xlab = "Skin Thickness", ylab = "Frequency")

hist(med_data$Insulin, main = "Distribution of Insulin Level", xlab = "Insulin Level", ylab = "Frequency")

hist(med_data$BMI, main = "Distribution of BMI", xlab = "BMI ", ylab = "Frequency")

hist(med_data$Pedigree, main = "Distribution of Pedigree", xlab = "Pedigree", ylab = "Frequency")

hist(med_data$Age, main = "Distribution of Age", xlab = "Age", ylab = "Frequency")

hist(med_data$Diagnosis, main = "Distribution of Diagnosis", xlab = "Daignosis Class", ylab = "Frequency")
```

### Find the best performing variables/features using a correlogram


```{r}
#Create a correlogram
# Load the required libraries


correlation_matrix <- cor(med_data[, c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI", "Pedigree", "Age", "Diagnosis")])

corrplot(correlation_matrix, method = "color", addCoef.col = "black")



# Find the correlation coefficients of the dependent variable (Diagnosis)
cor_with_diagnosis <- abs(correlation_matrix[,"Diagnosis"])

# Set a threshold for selecting the best features (e.g., absolute correlation > 0.3 or 0.5)
threshold <- 0.3

# Select the independent variables with correlation above the threshold
best_features <- names(cor_with_diagnosis[cor_with_diagnosis > threshold])

# Print the best features
print(best_features)

```

# Standardize your features to Gaussian distribution. Explain why it would be a good idea to standardize the features to Gaussian distribution

```{r}
#Extract the features
Pregnancies <- med_data$Pregnancies
Glucose <- med_data$Glucose
BloodPressure <- med_data$BloodPressure
SkinThickness <- med_data$SkinThickness
Insulin <- med_data$Insulin
BMI <-med_data$BMI
Pedigree <- med_data$Pedigree
Age <- med_data$Age


#Gaussaian Distribution

Pregnancies_n <- scale(Pregnancies)
Glucose_n <- scale(Glucose)
BloodPressure_n <- scale(BloodPressure)
SkinThickness_n <- scale(SkinThickness)
Insulin_n <- scale(Insulin)
BMI_n<- scale(BMI)
Pedigree_n <- scale(Pedigree)
Age_n <- scale(Age)



hist(Pregnancies_n, main = "Standardized Pregnancies", xlab = "Pregnancies ", ylab = "Frequency")
hist(Glucose_n, main = "Standardized Glucose", xlab = "Glucose level ", ylab = "Frequency")
hist(BloodPressure_n, main = "Standardized BloodPressure", xlab = "BloodPRessure ", ylab = "Frequency")
hist(SkinThickness_n, main = "Standardized SkinThickness", xlab = "SkinThickness", ylab = "Frequency")
hist(Insulin_n, main = "Standardized Insulin", xlab = "BMI ", ylab = "Frequency")
hist(BMI_n, main = "Standardized BMI", xlab = "BMI ", ylab = "Frequency")
hist(Pedigree_n, main = "Standardized Pedigree", xlab = "Pedigree", ylab = "Frequency")
hist(Age_n, main = "Standardized Age", xlab = "Age ", ylab = "Frequency")
```

# With MEDIAN

###Model1: BEST FEATURES with median Imputation

```{r}
#Create a data set with the standardized values
model_data <- data.frame(Pregnancies = Pregnancies_n, Glucose = Glucose_n, BloodPressure= BloodPressure_n, SkinThickness= SkinThickness_n , Insulin = Insulin_n, BMI = BMI_n, Pedigree = Pedigree_n,Age = Age_n, BMI = BMI_n, Diagnosis = med_data$Diagnosis)
#model_data


#Using a caret we will split out dataset into training and testing set
#install.packages("caret")
library(caret)

#Split the dataset
set.seed(123) #for reproducibility
train_indices1 <-createDataPartition(model_data$Diagnosis, p=0.8, list = FALSE)

train_data1 <- model_data[train_indices1,]
test_data1 <- model_data[-train_indices1,]
```

```{r}
# Check the linearity of the independent variables with dependent variables

plot(train_data1$Glucose, train_data1$Diagnosis, xlab = "Glucose", ylab = "Diagnosis", pch = 16, col = ifelse(train_data1$Diagnosis == 1, "red", "blue"))

plot(train_data1$BMI, train_data1$Diagnosis, xlab = "BMI", ylab = "Diagnosis", pch = 16, col = ifelse(train_data1$Diagnosis == 1, "red", "blue"))

```

```{r}
#Train model1 with best features which in our case is Glucose and BMI

model1 <- glm(Diagnosis ~ Glucose + BMI, data = train_data1, family = binomial)
model1

#create steps
step_model <- step(model1)

prediction1 <- predict(model1, new_data = test_data1, type = "response")
```

```{r}
# Assuming that the dependent variable is a factor, convert the predictions to binary class labels
prediction1 <- ifelse(prediction1 >= 0.5, 1, 0)

```


```{r}

par(mfrow = c(1,2))

pred_in1 <- predict(model1,newdatas = train_data1,type = "response")
prediction_in1 <- ifelse(pred_in1<0.29,0,1)
tab<-table(as.factor(train_data1$Diagnosis),prediction_in1)
roc.plot(train_data1$Diagnosis=="1", pred_in1)$roc.vol

```

```{r}

pred_out1 <- predict(model1,newdata = test_data1,type = "response")
prediction_out1 <- ifelse(pred_out1<0.29,0,1)
table(as.factor(test_data1$Diagnosis),prediction_out1)

```

```{r}
roc.plot(test_data1$Diagnosis=="1", pred_out1)$roc.vol
```


# Calculate model performance metrics


```{r}
# Convert predicted values to factors
prediction_out1 <- as.factor(ifelse(pred_out1 < 0.29, 0, 1))

# Convert reference labels to factors with the same levels
reference <- as.factor(test_data1$Diagnosis)
levels(reference) <- levels(prediction_out1)

# Calculate model performance metrics
library(caret)

# Accuracy
accuracy <- confusionMatrix(data = prediction_out1, reference = reference)$overall['Accuracy']

# F1 score
f1_score <- confusionMatrix(data = prediction_out1, reference = reference)$byClass['F1']

# Precision
precision <- confusionMatrix(data = prediction_out1, reference = reference)$byClass['Pos Pred Value']

#Recall
recall <- confusionMatrix(data= prediction_out1, reference = reference)$byClass['Sensitivity']

# Print the performance metrics
cat("Accuracy :", accuracy, "\n")
cat("F1 Score:", f1_score, "\n")
cat("Precision:", precision, "\n")
cat("Recall:", recall,"\n")

```


###Model2: ALL VARIABLES

```{r}
#Create a data set with best features and diagnosis
model_data2 <- data.frame(Pregnancies = Pregnancies_n, Glucose = Glucose_n, BloodPressure = BloodPressure_n, Insulin=Insulin_n, SkinThickness = SkinThickness_n, Pedigree = Pedigree_n, BMI = BMI_n, Age = Age_n , Diagnosis = med_data$Diagnosis)
model_data2

#Using a caret we will split out dataset into training and testing set
#install.packages("caret")
library(caret)

#Split the dataset
set.seed(123) #for reproducibility
train_indices2 <-createDataPartition(model_data2$Diagnosis, p=0.8, list = FALSE)

train_data2<- model_data2[train_indices2,]
test_data2 <- model_data2[-train_indices2,]
```
```{r}
#Train model

model2 <- glm(Diagnosis ~ Pregnancies + Glucose + SkinThickness + BloodPressure + Insulin + Pedigree + BMI + Age, data = train_data2, family = binomial)
model2

#step_model <- step(model.glm)

prediction2 <- predict(model2, new_data = test_data2, type = "response")
```
```{r}
# Assuming that the dependent variable is a factor, convert the predictions to binary class labels
prediction2 <- ifelse(prediction2 >= 0.5, 1, 0)



```

# Model2 with Median Evaluation


```{r}

par(mfrow = c(1,2))

pred_in2 <- predict(model2,newdatas = train_data2,type = "response")
prediction_in2 <- ifelse(pred_in2<0.29,0,1)
tab<-table(as.factor(train_data2$Diagnosis),prediction_in2)
roc.plot(train_data2$Diagnosis=="1", pred_in2)$roc.vol

```

```{r}

pred_out2 <- predict(model2,newdata = test_data2,type = "response")
prediction_out2 <- ifelse(pred_out2<0.29,0,1)
table(as.factor(test_data2$Diagnosis),prediction_out2)

```

```{r}
roc.plot(test_data2$Diagnosis=="1", pred_out2)$roc.vol

```


# Calculate model performance metrics

```{r}
# Convert predicted values to factors
prediction_out2<- as.factor(ifelse(pred_out2 < 0.29, 0, 1))

# Convert reference labels to factors with the same levels
reference <- as.factor(test_data2$Diagnosis)
levels(reference) <- levels(prediction_out2)

# Calculate model performance metrics
library(caret)

# Accuracy
accuracy <- confusionMatrix(data = prediction_out2, reference = reference)$overall['Accuracy']

# F1 score
f1_score <- confusionMatrix(data = prediction_out2, reference = reference)$byClass['F1']

# Precision
precision <- confusionMatrix(data = prediction_out2, reference = reference)$byClass['Pos Pred Value']

#Recall
recall <- confusionMatrix(data= prediction_out2, reference = reference)$byClass['Sensitivity']

# Print the performance metrics
cat("Accuracy :", accuracy, "\n")
cat("F1 Score:", f1_score, "\n")
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")

```
