---
title: "Prj_trail_1_cont"
author: "Shourav Dey"
date: "2023-06-03"
output: html_document
---

```{r}
getwd()
```


Importing the data

```{r}
data <-  read.csv("patients.csv", header=TRUE, sep = ",",)
data
```

```{r}
View(data)
```

```{r}
summary(data)
```


# Count the number of "0" values in each column
```{r}
zero_counts <- colSums(data == 0, na.rm = TRUE)

zero_counts
```

#--------------------------BMI-------------------------#

 Underweight: BMI below 18.5
 Normal weight: BMI between 18.5 and 24.9
 Overweight: BMI between 25 and 29.9
 Obesity: BMI 30 or higher
#-----------------------------------------------------------

#---------***** Before Imputation********--------------

```{r}
library(dplyr)

data_summary_BMI <- data %>%
  summarize(mean_value = mean(BMI),
            sd_value = sd(BMI),
            iqr_value = IQR(BMI),
            median_value = median(BMI))

data_summary_BMI

```

# Findings

mean_value        sd_value     iqr_value
31.99258           7.88416      9.3
median BMI 32


# 11 missing values

```{r}
fat_count <- colSums(data >= 32, na.rm = TRUE)

fat_count

```

# Need to do fat count again to see how many values we have over the mean




```{r}
library(ggplot2)

```
##### HISTOGRAM before imputation

# Better visualization


```{r}
hist_bmi <- ggplot(data) +
  geom_histogram(aes(x = BMI, y = ..density..), fill = "orange", color = "black", binwidth = 0.2) +
  geom_density(aes(x = BMI, color = "KDE"), bw = 0.1) +
  theme(panel.background = element_rect(fill = "gray90"))

hist_bmi

```
There are some good amount of outliers present
 There are 11 counts of 0 values present which is not possible
 Both the mean and the median value suggests the dataset is filled with Obese people
 The data with BMI more than 45 are outliers or extreme cases, shouldn't be ignored
 THe shape however is bell shaped



# Boxplot before Imputation

```{r}
box_bmi_trial <- ggplot(data, aes(x=BMI))+geom_boxplot()

box_bmi_trial

```
 The box plot varifies that data with BMI > 45 are outliers
 the data with 0 are outliers
 But I don't think we should ignore


# --BloodPressure

```{r}
data_summary_bp <- data %>%
  summarize(mean_value = mean(BloodPressure),
            sd_value = sd(BloodPressure),
            iqr_value = IQR(BloodPressure),
            median_value = median(BloodPressure))

data_summary_bp

```
 mean_value    sd_value    iqr_value       median_value
 69.10547      19.35581     18                   72

# According to AHA
# Normal blood pressure:  Diastolic <80 mm Hg
# Elevated blood pressure: Diastolic <80 mm Hg
# Hypertension Stage 1:  Diastolic 80-89 mm Hg
# Hypertension Stage 2: Diastolic ≥90 mm Hg

Median value lies on the Elevated blood pressure
Mean suggest normal bp

# 35 missing values


#----------Histogram of original BP data-------------


```{r}
bp_hist_trial <- ggplot(data)+ geom_histogram(aes(x=BloodPressure, y = ..density..), fill = "orange", color ="black", binwidth =0.2)+
                  geom_density(aes(x=BloodPressure, color ="KDE"), bw=0.1)+
                  theme(panel.background = element_rect(fill ="gray90"))

bp_hist_trial

```
# Heavy density of 0 outliers
# However, the bell shape is pretty obvious representing normal distribution in the dataset
# Presence of some minor outliers on both extremes but nothing extreme



#----------Boxplot for BP w/o imputation--------------

```{r}
bp_box_trial <- ggplot(data, aes(x=BloodPressure))+geom_boxplot()

bp_box_trial

```




#------------------------ Rough trial (most likely useles)------------#

#--- Checking correlation between BMI and BP


library(ggplot2)

```{r}
bmi_bp_scatter <- ggplot(data, aes(x = BMI, y = BloodPressure)) +
  geom_point()

bmi_bp_scatter

```

```{r}
corr_bmi_bp <- cor(data$BMI, data$BloodPressure)

corr_bmi_bp

```



#------------------------------------------------------------------------#


#--------------Corr BMI vs Diagnosis---------

#Bar plot

```{r}
ggplot(data, aes(x=Diagnosis, y=BMI)) +
  geom_bar(stat = "summary", fun="mean")

```
#Violin plot-----------Useless

```{r}
ggplot(data, aes(x = Diagnosis, y = BMI)) +
  geom_violin()
```

#Box plot--------Useless

ggplot(data, aes(x = Diagnosis, y = BMI)) +
  geom_boxplot()

#Scatter plot------ Seems like a vague relationship

ggplot(data, aes(x=BMI, y=as.numeric(Diagnosis)))+
  geom_point()


#correlation

corr_bmi_diag <- cor(data$BMI, data$Diagnosis)

corr_bmi_diag
# 
# corr_bmi_diag
#  0.2926947


# From Jass

 Insulin
 Normal 80-120
 Mean 127.8
 insulin and glucose are related
 if you are 10 points above 120



```{r}

max_zeros <- max(rowSums(data == 0))
rows_with_max_zeros <- which(rowSums(data == 0) == max_zeros)

# Print the rows with maximum zeros
print(data[rows_with_max_zeros, ])

```

# New data frame with diagnosis 0 & 1


```{r}
# Creating a dataframe with diagnosis = 0
data_diag_0 <- data[data$Diagnosis == 0, ]

# Creating a dataframe with diagnosis = 1
data_diag_1 <- data[data$Diagnosis == 1, ]

```

500 Not diagnosed
268 Diagnosed

```{r}

head(data_diag_0)

```




```{r}


# Assuming your dataframe is called 'data'
# Create a vector of the mentioned variables
variables <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI", "Pedigree")

# Filter the rows based on the conditions
filtered_data <- data[rowSums(data[variables] == 0) > 2, ]

# View the resulting dataframe
filtered_data


```




```{r}

View(filtered_data)


```



#------------------Histogram and counts of Features before Imputation----------------#



#-----------Glucose------------------

# Histogram of entire glucose
```{r}
data %>%
  ggplot(aes(x = Glucose)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Glucose Level ",
       y = "Number of Women")+ggtitle("Variation of glucose in women")


```
# THe Hisotgram seems skewed to the left. There are presence of 


#Density and Box plot of Not Diagnosed


```{r}


library(gridExtra)

p2 <- ggplot(data_diag_0, aes(x = Glucose)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Glucose", y = "Density", title = "Density plot of glucose")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = Glucose)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Variation of glucose in women")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

# Boxplot and Density plot of Diagnosed

```{r}

library(gridExtra)

p2 <- ggplot(data_diag_1, aes(x = Glucose)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Glucose", y = "Density", title = "Density plot of glucose")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = Glucose)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Variation of glucose in women")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```
```{r}
summary(data)
```

```{r}
summary(data_diag_0)
```

```{r}
summary(data_diag_1)

```



#----------------------SkinThickness-------------------------


```{r}

data %>%
  ggplot(aes(x = SkinThickness)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "SkinThickness ",
       y = "Number of Women")+ggtitle("Variation in Skinthickness in women")


```
# Normally distributed, but heavy presence of outliers


------------No Diabetes


```{r}
library(gridExtra)

p2 <- ggplot(data_diag_0, aes(x = SkinThickness)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "SkinThickness", y = "Density", title = "SkinThickness  NO Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = SkinThickness)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Skinthickness No Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

-------------------DIABETES

```{r}

library(gridExtra)

p2 <- ggplot(data_diag_1, aes(x = SkinThickness)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "SkinThickness", y = "Density", title = "Diabetes vs SkinThickness")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = SkinThickness)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Diabetes vs SkinThickness")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```



#---------------------BMI___________________



```{r}
data %>%
  ggplot(aes(x = BMI)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "BMI ",
       y = "Number of Women")+ggtitle("Variation of BMI in data")

```
skewed to the right

---------------- Diabetes


```{r}
library(gridExtra)

p2 <- ggplot(data_diag_1, aes(x = BMI)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "BMI", y = "Density", title = "BMI vs Diabetes")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = BMI)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("BMI Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```






---------BMI vs No Diabetes

```{r}

library(gridExtra)

p2 <- ggplot(data_diag_0, aes(x = BMI)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "BMI", y = "Density", title = "Density plot BMI without Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = BMI)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("BMI without Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)


```



```{r}
patientdata = read.csv("patients.csv", header=TRUE, sep = ",",)

```


# ---------------------BloodPressure---------------

```{r}
data %>%
  ggplot(aes(x = BloodPressure)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Blood Pressure ",
       y = "Number of Women") +
ggtitle("Blood Pressure in entire dataset")


```

# Normally distribute with slight skewness
# Normality (line) test will be a good test
# Q-q density plot

-------------------------------------Diabetes and BP----------------
```{r}
library(gridExtra)

p2 <- ggplot(data_diag_1, aes(x = BloodPressure)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "BMI", y = "Density", title = "BP vs Diabetes")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = BloodPressure)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Blood Pressure Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


----------------BP No Diabetes-----------------------------

```{r}

library(gridExtra)

p2 <- ggplot(data_diag_0, aes(x = BloodPressure)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Bloodpressure", y = "Density", title = "BP NO Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = BloodPressure)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Blood Pressure No Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


# ------------------------Insulin---------------------

```{r}

data %>%
  ggplot(aes(x = Insulin)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Insulin ",
       y = "Number of Women") + ggtitle("Insulin in entire dataset")

```
# Completely obselete dataset, Need Severe care.
# Heavy on the outliers
# WHatever distribution is remaining is right skewed.
# Seems presence of outliers on the extreme ends

--------------------------- Insulin vs Diabetes

```{r}

p2 <- ggplot(data_diag_1, aes(x = Insulin)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Insulin", y = "Density", title = "Insulin vs Diabetes")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = Insulin)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Insulin & Diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)



```
# Unsettling distribution and presence of outliers on the extreme ends are visible


```{r}


p2 <- ggplot(data_diag_0, aes(x = Insulin)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Insulin", y = "Density", title = "Insulin NO Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = Insulin)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Insulin level NO diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)



```
# Same .. Needs extreme care


-----------------------------------Pedigree-----------------------

```{r}
data %>%
  ggplot(aes(x = Pedigree)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Pedigree ",
       y = "Number of Women") + ggtitle("Pedigree in entire dataset")

```
# Somewhat acceptable however presence of outliers in the extreme end.
# Furthermore, the data is skewed to the right

---------------------------------Pedigree & Diabetes

```{r}

p2 <- ggplot(data_diag_1, aes(x = Pedigree)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Pedigree", y = "Density", title = "Pedigree & Diabetes")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = Pedigree)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Pedigree & diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)



```



---------------Pedigree NO Diabetes----------------------


```{r}
p2 <- ggplot(data_diag_0, aes(x = Pedigree)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Pedigree", y = "Density", title = "Pedigree NO Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = Pedigree)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Pedigree NO diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)


```

#Distribution is okay but heavily right skewed. Therefore presence of outliers in the extreme boundaries, evident from the box plot as well



#---------------Pregnancies----------------

```{r}
data %>%
  ggplot(aes(x = Pregnancies)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Pregnancies ",
       y = "Number of Women") + ggtitle("Pregnancies in entire dataset")


```

```{r}

# Count the occurrences of each unique value in the 'Pregnancies' column
pregnancies_counts <- table(data$Pregnancies)

# Create a bar plot
ggplot(data = data.frame(Pregnancies = names(pregnancies_counts),
                         Count = as.vector(pregnancies_counts))) +
  geom_bar(aes(x = Pregnancies, y = Count), stat = "identity") +
  xlab("Pregnancies") +
  ylab("Count") +
  ggtitle("Count of Pregnancies")

```

```{r}
# Create a box plot
ggplot(data, aes(x = "", y = Pregnancies)) +
  geom_boxplot() +
  xlab("") +
  ylab("Pregnancies") +
  ggtitle("Box Plot of Pregnancies")

```
the dataset is such that it does not take the zeroes as outliers at all. Therefore other methods of imputation are required.

```{r}

ggplot(data, aes(x = Pregnancies)) +
  geom_density(fill = "skyblue", alpha = 0.7) +
  xlab("Pregnancies") +
  ylab("Density") +
  ggtitle("Density Plot of Pregnancies")

```
The column itself is heavily right skewed



# ------------ Age--------------


```{r}

age_counts <- table(data$Age)

# Create a bar plot
ggplot(data = data.frame(Age = names(age_counts),
                         Count = as.vector(age_counts))) +
  geom_bar(aes(x = Age, y = Count), stat = "identity") +
  xlab("Age") +
  ylab("Count") +
  ggtitle("Visualization for age")

```

```{r}

data %>%
  ggplot(aes(x = Age)) +
  geom_histogram(fill = "Orange", color = "black") +
  theme_minimal() +
  labs(x = "Age ",
       y = "Number of Women") + ggtitle("Histogram Visualization for Age")

```

No zero counts.Presence of Outliers in the 80 ish side.
Distribution is heavily skewed to the right.




--------------Age & Diagbetes

```{r}

p2 <- ggplot(data_diag_1, aes(x = Age)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Age", y = "Density", title = "Age & Diabetes")

p1 <- ggplot(data_diag_1, aes(x = Diagnosis, y = Age)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Age & diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)


```



```{r}

p2 <- ggplot(data_diag_0, aes(x = Age)) +
  geom_density(alpha = 0.8, fill = "Orange", color = "black") +
  theme(legend.position = "bottom") +
  labs(x = "Age", y = "Density", title = "Age NO Diabetes")

p1 <- ggplot(data_diag_0, aes(x = Diagnosis, y = Age)) +
  geom_boxplot() +
  theme(legend.position = "bottom") +
  ggtitle("Age NO diabetes")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


--------------Diagnosis-------------

```{r}

# Create a stacked bar plot
ggplot(data, aes(x = factor(Diagnosis), fill = factor(Diagnosis))) +
  geom_bar() +
  xlab("Diagnosis") +
  ylab("Count") +
  ggtitle("Stacked Bar Plot of Diagnosis")

```














#-------------- KNN -----------------------
 Handling the consecutive zeros in the rows

```{r}


#  a vector of the variables

variables <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI", "Pedigree")

# Filter the rows based on the conditions
filtered_data <- patientdata[rowSums(patientdata[variables] == 0) > 2, ]

# View the resulting dataframe
filtered_data

```

```{r}
#following block of code deletes the rows which has more than 2 zeros in it.#
patientdata <- patientdata[rowSums(patientdata[variables] == 0) <= 2, ]

```

```{r} 
summary(patientdata)
```

```{r}
#Counts of zeros in each column
zero_counts <- colSums(patientdata == 0)
print(zero_counts)

```


View(patientdata)


```{r}
#Counts of zeros in each column
zero_counts <- colSums(patientdata == 0)
print(zero_counts)

```
```{r}
summary(patientdata)
```
```{r}
#replacing the zeros with NA

 

# Specify the columns to exclude (1, 8, 9)
columns_to_exclude <- c(1, 8, 9)

 

# Replace zeros with NA in the remaining columns
patientdata[, -columns_to_exclude][patientdata[, -columns_to_exclude] == 0] <- NA

``` 

```{r}
# View the updated dataset
summary(patientdata)

```

```{r}
install.packages("VIM")

```
 
```{r}
#imputing the zeros with KNN
# Specify the columns to impute (excluding columns pregnancy, Age, and Diagnosis)
columns_to_impute <- setdiff(seq_along(patientdata), c(1, 8, 9))

```
 
```{r}
# Perform kNN imputation on the selected columns
patientdata[, columns_to_impute] <- kNN(patientdata[, columns_to_impute], k = 5)

```
```{r} 

# View the updated dataset
patientdata

```
```{r}

summary(patientdata)

```

```{r}

zero_counts <- colSums(patientdata == 0)
print(zero_counts)

```















